---
title: "Final Project"
format: pdf
editor: visual
---

```{r}
#| label: load_packages
# Load necessary packages ----
library(ggplot2)
library(dplyr)
library(knitr)
library(tinytex)
library(tidyverse)
# Google R style guide used
```

```{r}
#| label: load_data
student_raw <-  read.csv(
  "https://raw.githubusercontent.com/Stat184-Spring2025/Sec4_FP_JonasP_WillB_AlexH/refs/heads/main/student_lifestyle_dataset..csv",
  sep = ","
)
```

```{r}
#| label: GPA to 4.0 scale
student_4 <- student_raw %>%
  mutate(Grades = Grades * (4/10)) 
```

## Introduction

## Exploratory data Analyis

### Frequency table

### Frequency graphs

## Main Analysis

### Hypothesis

### Data analysis

### Further research? (hypothesis testing, regression, ANOVA, or machine learning)

\

### Supporting research

### Conclusion

## Data Information

### Main Data Set

1.  Data Tidying

2.  Data Wrangling

3.  Data Cleaning

### Secondary Data Set

1.  Data Tidying

2.  Data Wrangling

3.  Data Cleaning

### Fair and Care Principles

## Open Science process

## Code Apendix

## Citations
